Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters

### 2019-08-31 paper.tex

\begin{abstract}
    Big Data has become prominent throughout many scientific fields and, as
    a result, scientific communities have sought out Big Data frameworks to
    accelerate the processing of their increasingly data-intensive
    pipelines. However, while scientific communities typically rely on
    High-Performance Computing (HPC) clusters for the parallelization of
    their pipelines, many popular Big Data frameworks such as Hadoop and
    Apache Spark were primarily designed to be executed on dedicated
    commodity infrastructures. This paper evaluates the benefits of pilot
    jobs over traditional batch submission for Apache Spark on HPC
    clusters. Surprisingly, our results show that the speed-up provided by
    pilot jobs over batch scheduling is moderate to non-existent (0.98 on
    average) despite the presence of long queuing times. In addition, pilot
    jobs provide an extra layer of scheduling that complicates debugging
    and deployment. We conclude that traditional batch scheduling should
    remain the default strategy to deploy Apache Spark applications on
    HPC clusters.

@@

    With HPC clusters, client mode is not necessarily possible on the login node as the
    driver client may require more resources than permitted. Moreover, due to cluster
    security policies, it may be difficult or impossible, to execute the driver from
    an external computer, such as a personal laptop. Therefore, for both batch and pilot
    scheduling, it is necessary to launch the driver from within the Slurm jobs. That is, in cluster mode.

    Spark also provides master fault tolerance through single-node recovery, wherever a shared
    filesystem is available. To enable this, a recovery folder must be set. Should the master
    fail, a new master will be able to takeover based on the information available in the recovery
    folder.

@@

    Assuming a divisible load, i.e., the application can be divided in any
    number of tasks, the makespan can be written using the following
    expression, which holds for both batch and pilot execution modes:
    \begin{equation}
        M = \frac{C}{W} \label{eq:mcw}
    \end{equation}
    where:
    \begin{itemize}
        \item $M$ is the makespan of the application
        \item $C$ is the total wall-time of the application
        \item $W$ is the \emph{average} number of Spark workers throughout the execution
    \end{itemize}
    The average number of workers $W$ allows the model to take into account
     variable queuing times. It is computed as follows:
    \begin{equation}
        W = \frac{1}{M}\int_0^M{w(t)dt}\label{eq:avgw}
    \end{equation}
    where $w(t)$ is the number of workers available at time $t$. When the
    application is not subject to any scheduling or queuing time, the
    average number of workers equals the number of workers requested.

    Therefore, assuming a fixed total wall-time, the relation
    between batch and pilot jobs can be represented as:
    \begin{equation}
        \frac{M_{batch}}{M_{pilot}} = \frac{W_{pilot}}{W_{batch}}\label{eq:makespancomp}
    \end{equation}
    where:
    \begin{itemize}
        \item $M_{batch}$ is the makespan of the batch application
        \item $M_{pilot}$ is the makespan of the pilot application
        \item $W_{pilot}$ is the average number of workers of the pilot application
        \item $W_{batch}$ is the average number of workers of the batch application
    \end{itemize}
    We will use this relation to discuss our results later on. It
    corresponds to an ideal case where no data or other overhead is
    present: only queuing times are included, through the integration in
    Equation~\ref{eq:avgw}.

@@

It was also found that having 16 pilots was generally slower than 8 for the same
configuration. This was particularly the case on B\'eluga. We suspect this might be
related to the number of Slurm jobs permitted for a user on the backfill queue. For
all configurations, we ran batch, 8 pilots and 16 pilots concurrently. This would have
resulted in 25 Slurm jobs trying the access the queue concurrently. On
B\'eluga, the max number of jobs in the backfill queue is configured to a
total of 10 jobs. For 16 nodes, this would at best be only 62.5\% of the
pilots being scheduled at a given time if all resources are available.
The next 6 jobs would then have to wait 3 minutes before having an opportunity to enter the
backfill. Cedar, on the other hand, permits 40 user jobs to be backfilled at a given time.
Therefore, on Cedar, all concurrent Slurm jobs had an opportunity to be backfilled at
the same time, which may explain why this behaviour is more apparent on B\'eluga. Furthermore,
while both pilot configurations could be placed in B\'eluga's low memory nodes, batch requests
had to be place on B\'eluga's medium memory nodes. B\'eluga has 172 low memory nodes and
516 mid memory nodes. As pilots would get priority for the low memory nodes first and 
low memory nodes are less frequent that medium memory nodes, it is possible that this
may have increased the overall queuing times of the pilots. However, Cedar's basic
node has enough memory to for all batch and pilot configuration, therefore, this could not
have affected queuing times in Cedar.

@@

The motivation to use pilots, however, is dynamic scheduling on HPC clusters. While we
investigated the queuing time differences with respect to differences in available resources,
the wall-times were kept static. Had wall-time been underestimated to the point where the Slurm jobs
would terminate prior to application completion, pilots might have been preferable to batch
as batch scheduling has no mechanism to restart. Even if such a mechanism was in place, the entire
cluster would be shutdown and need to restart, which has additional overheads. With pilots, as long as 
there is at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
Spark provides built-in fault-tolerance for not only the workers, but the master and driver as well. However,
it is also likely that with faster queuing times, all resources would be allocated at the same time. If wall-time
were to be underestimated, in this case, checkpointing and restarting the entire application from the last
checkpoint would be the only option. This too has limitations as the last checkpoint may not be recent,
necessitating significant amounts of recomputation. It
might even be desirable to have many small resource requests in such a scenario, as it becomes less likely for all pilots to start at once as they become more numerous, as can be seen with 16 pilots when compared to 8. This would allow
for more chances for the application to reach completion without failing entirely.

Many failures can be found, particularly with regards to the pilots. Most
of these have been a result of Spark failing to start properly. The driver
either would not return a submission ID after being started, or would
remain in \texttt{SUBMITTED} status until the wall-time expiration, having
not processed any data. As these errors become fewer with increased
resources, it is possible they are entirely related to lack of resources.
Furthermore, it may also explain why batch submission is more frequently successful. Not
only are all the workers that were requested available when the pilot
begins running, but batch also has one extra executor available thanks to the driver
process running directly on the process which had created it, rather than on a
worker process. Conversely, it is not only the pilot applications that
fail, the batch applications do experience some failures as well. It can be
seen in Figure~\ref{fig:makespansbeluga} that the batch application failed
twice: once in Configuration~2 and the other in Configuration~4. In both
cases, the reason for failure was the same: lack of available resources
despite all executors having been assigned. These failures might be due to
running on a shared HPC cluster. Either the resources were down, or other users might
have been using them, resulting in the resources being unavailable to our
application.

>>>>

\begin{abstract}
    Big Data has become prominent throughout many scientific fields, and as
    a result, scientific communities have sought out Big Data frameworks to
    accelerate the processing of their increasingly data-intensive
    pipelines. However, while scientific communities typically rely on
    High-Performance Computing (HPC) clusters for the parallelization of
    their pipelines, many popular Big Data frameworks such as Hadoop and
    Apache Spark were primarily designed to be executed on dedicated
    commodity infrastructures. This paper evaluates the benefits of pilot
    jobs over traditional batch submission for Apache Spark on HPC
    clusters. Surprisingly, our results show that the speed-up provided by
    pilot jobs over batch scheduling is moderate to non-existent (0.98 on
    average) despite the presence of long queuing times. In addition, pilot
    jobs provide an extra layer of scheduling that complicates debugging
    and deployment. We conclude that traditional batch scheduling should
    remain the default strategy to deploy Apache Spark applications on
    HPC clusters.

@@

    With HPC clusters, client mode is not necessarily possible on the login node as the
    driver client may require more resources than permitted. Moreover, due to cluster
    security policies, it may be difficult or impossible to execute the driver from
    an external computer, such as a personal laptop. Therefore, for both batch and pilot
    scheduling, it is necessary to launch the driver from within the Slurm jobs. That is, in cluster mode.

    Spark also provides master fault tolerance through single-node recovery wherever a shared
    filesystem is available. To enable this, a recovery folder must be set. Should the master
    fail, a new master will be able to takeover based on the information available in the recovery
    folder.

@@

    Assuming a divisible load, i.e., the application can be divided in any
    number of tasks, the makespan can be written using the following
    expression, which holds for both batch and pilot execution modes:
    \begin{equation}
        M = \frac{C}{W} \label{eq:mcw}
    \end{equation}
    where:
    \begin{itemize}
        \item $M$ is the makespan of the application
        \item $C$ is the total execution time of the application
        \item $W$ is the \emph{average} number of Spark workers throughout the execution
    \end{itemize}
    The average number of workers $W$ allows the model to take into account
     variable queuing times. It is computed as follows:
    \begin{equation}
        W = \frac{1}{M}\int_0^M{w(t)dt}\label{eq:avgw}
    \end{equation}
    where $w(t)$ is the number of workers available at time $t$. When the
    application is not subject to any scheduling or queuing time, the
    average number of workers equals the number of workers requested.

    Therefore, assuming a fixed total execution time, the relation
    between batch and pilot jobs can be represented as:
    \begin{equation}
        \frac{M_{batch}}{M_{pilot}} = \frac{W_{pilot}}{W_{batch}}\label{eq:makespancomp}
    \end{equation}
    where:
    \begin{itemize}
        \item $M_{batch}$ is the makespan of the batch application
        \item $M_{pilot}$ is the makespan of the pilot application
        \item $W_{pilot}$ is the average number of workers of the pilot application
        \item $W_{batch}$ is the average number of workers of the batch application
    \end{itemize}
    We will use this relation to discuss our results later on. It
    corresponds to an ideal case where no data or other overhead is
    present: only queuing times are included, through the integration in
    Equation~\ref{eq:avgw}.

@@

It was also found that having 16 pilots was generally slower than 8 for the same
configuration. This was particularly the case on B\'eluga. We suspect this might be
related to the number of Slurm jobs permitted for a user on the backfill queue. For
all configurations, we ran batch, 8 pilots and 16 pilots concurrently. This would have
resulted in 25 Slurm jobs trying the access the queue concurrently. On
B\'eluga, the max number of jobs in the backfill queue is configured to a
total of 10 jobs. For 16 nodes, this would at best be only 62.5\% of the
pilots being scheduled at a given time if all resources are available.
The next 6 jobs would then have to wait 3 minutes before having an opportunity to enter the
backfill. Cedar, on the other hand, permits 40 user jobs to be backfilled at a given time.
Therefore, on Cedar, all concurrent Slurm jobs had an opportunity to be backfilled at
the same time, which may explain why this behaviour is more apparent on B\'eluga. Furthermore,
while both pilot configurations could be placed in B\'eluga's low memory nodes, batch requests
had to be placed on B\'eluga's medium memory nodes. B\'eluga has 172 low memory nodes and
516 mid memory nodes. As pilots would get priority for the low memory nodes first and 
low memory nodes are less frequent that medium memory nodes, it is possible that this
may have increased the overall queuing times of the pilots. However, Cedar's basic
node has enough memory to for all batch and pilot configuration, therefore, this could not
have affected queuing times in Cedar.

@@

The motivation to use pilots, however, is dynamic scheduling on HPC clusters. While we
investigated the queuing time differences with respect to differences in available resources,
the wall-times were kept static. Had wall-times been underestimated to the point where the Slurm jobs
would terminate prior to application completion, pilots might have been preferable to batch
as batch scheduling has no mechanism to restart. Even if such a mechanism was in place, the entire
cluster would be shutdown and need to restart, which has additional overheads. With pilots, as long as 
there is at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
Spark provides built-in fault-tolerance for not only the workers, but the master and driver as well. However,
it is also likely that with faster queuing times, all resources would be allocated at the same time. If wall-times
were to be underestimated, in this case, checkpointing and restarting the entire application from the last
checkpoint would be the only option. This too has limitations as the last checkpoint may not be recent,
necessitating significant amounts of recomputation. It
might even be desirable to have many small resource requests in such a scenario, as it becomes less likely for all pilots to start at once as they become more numerous, as can be seen with 16 pilots when compared to 8. This would allow
for more chances for the application to reach completion without failing entirely.

Many failures can be found, particularly with regards to the pilots. Most
of these have been a result of Spark failing to start properly. The driver
either would not return a submission ID after being started, or would
remain in \texttt{SUBMITTED} status until the wall-time expiration, having
not processed any data. As these errors become fewer with increased
resources, it is possible they are entirely related to lack of resources.
Furthermore, it may also explain why batch submission is more frequently successful. Not
only are all the workers that were requested available when the pilot
begins running, but batch also has one extra executor available thanks to the driver
process running directly on the process which had created it, rather than on a
worker process. Conversely, it is not only the pilot applications that
fail; batch applications do experience some failures as well. It can be
seen in Figure~\ref{fig:makespansbeluga} that the batch application failed
twice: once in Configuration~2 and the other in Configuration~4. In both
cases, the reason for failure was the same: lack of available resources
despite all executors having been assigned. These failures might be due to
running on a shared HPC cluster. Either the resources were down, or other users might
have been using them, resulting in the resources being unavailable to our
application.
